Nitrogen's form determines its fate in the North Pacific Subtropical Gyre during PARAGON I

### 2025-05-17 manuscript/manuscript.Rmd
Rerunning manuscript figures with fixed metadata

```{r load data, eval=FALSE}
parametadata <- read_csv("metadata/parametadata.csv") %>%
  # filter(IS_type=="noIS") %>%
  filter(polarity=="pos") %>%
  filter(samp_type=="Smp") %>%
  mutate(startime=factor(startime, levels=c("Morn", "Eve"))) %>%
  mutate(depth=factor(depth, levels=c("Surf", "Deep"))) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Nit", "Arg", "GMP"))) %>%
  arrange(amendment, depth, timepoint_num, startime) %>%
  mutate(timepoint=fct_inorder(timepoint))

@@

```{r}
xamend_metab_gp <- labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, timepoint, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(depth=="Surf") %>%
  mutate(compound_name=str_remove_all(compound_name, "^L-")) %>%
  mutate(compound_name=str_remove(compound_name, "/beta-Glutamic acid")) %>%
  mutate(compound_class=case_when(
    compound_name%in%c("Glutamine", "Aspartic acid", "Asparagine", "Glutamic acid", "Ectoine", "Theanine?", "Citrulline", "Guanine", "Guanosine", "Arginine") ~ compound_name,
    compound_name%in%maa_names~"MAAs",
    compound_name%in%aa_names~"Other amino acids",
    compound_name%in%nucleo_names~"Other nucleic acids",
    compound_name%in%betaine_names~"Betaines",
    TRUE~"All others"
  )) %>%
  mutate(compound_class=factor(compound_class, levels=class_levels)) %>%
  summarise(avg_frac=mean(labeled_frac), .by = c(compound_class, amendment, timepoint, tripl)) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Nit", "Arg", "GMP"),
                          labels=c("+Ammonia", "+Nitrate", "+Arginine", "+GMP"))) %>%
  ggplot() +
  geom_raster(aes(x=timepoint, y=tripl, fill=avg_frac)) +
  geom_vline(xintercept = c(1.5:4.5), color="grey80") +
  facet_nested(compound_class~amendment, scales = "free_x", switch="y") +
  scale_fill_gradientn(colours = viridis::viridis(10, alpha = seq(0, 1, length.out=10), direction = -1),
                       limits=c(0, 1), breaks=c(0, 0.25, 0.5, 0.75, 1),
                       labels=c("Fully unlabeled", "", "50% labeled", "", "Fully labeled")) +
  scale_x_discrete(expand = expansion()) +
  scale_y_discrete(expand = expansion()) +
  labs(x=NULL, y=NULL, fill=NULL) +
  theme_bw() +
  theme(panel.spacing.x=unit(0.5, "lines"), 
        panel.spacing.y = unit(0, "lines"),
        axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle=0, hjust = 1),
        strip.background.y = element_rect(fill=NA, color=NA),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "top", 
        legend.key.width = unit(1, "null")) +
  guides(
    fill=guide_colorbar(theme=theme(
      legend.frame = element_rect(color = "black"),
      legend.ticks = element_line(color="black"),
      legend.text = ggtext::element_markdown(hjust=c(0, 0, 0.5, 1, 1)))
    )
  )
ggsave("xamend_metab_gp.png",plot = xamend_metab_gp, device = "png", path = "manuscript/figures",
       width = 6.5, height = 5.5, units = "in", dpi = 300)
```

@@

```{r fig surface vs deep labeling rates}
surfdeep_demo_cmpd_gp <- labeled_fracs %>%
  mutate(compound_name=str_remove_all(compound_name, "^L-")) %>%
  mutate(compound_name=str_remove(compound_name, "/beta-Glutamic acid")) %>%
  mutate(compound_class=case_when(
    compound_name%in%c("Glutamine", "Aspartic acid", "Asparagine", "Glutamic acid", "Ectoine", "Theanine?", "Citrulline", "Guanine", "Guanosine", "Arginine") ~ compound_name,
    compound_name%in%aa_names~"Other amino acids",
    compound_name%in%nucleo_names~"Other nucleic acids",
    compound_name%in%betaine_names~"Betaines",
    TRUE~"All others"
  )) %>%
  mutate(compound_class=factor(compound_class, levels=setdiff(class_levels, "MAAs"))) %>%
  left_join(parametadata %>% distinct(shortname, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(compound_name=="Glutamic acid" & amendment=="Amm" |
           compound_name=="Arginine" & amendment=="Arg" |
           compound_name=="Guanosine" & amendment=="GMP") %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Arg", "GMP"),
                          labels=c("+Ammonia", "+Arginine", "+GMP"))) %>%
  select(-shortname) %>%
  summarise(mean_label=mean(labeled_frac), .by=c(compound_class, amendment, timepoint_num, depth)) %>%
  pivot_wider(names_from=depth, values_from=mean_label) %>%
  ggplot(aes(x=timepoint_num, color=compound_class, shape=compound_class)) +
  geom_segment(aes(xend=timepoint_num, y=Surf, yend=Deep), lwd=1) +
  geom_point(aes(y=Surf), fill="grey80", size=3, stroke=1) +
  geom_point(aes(y=Deep), fill="grey20", size=3, stroke=1) +
  stat_summary(aes(y=Surf, label=paste0(round(Surf-Deep, 2)*100, "%"), vjust=-sign(Surf-Deep)*1.5+0.75), 
               geom = "text", fun = mean) +
  facet_wrap(~amendment, nrow=1) +
  scale_color_discrete(drop=FALSE) +
  scale_x_log10(breaks=c(0.3, 3, 30), labels=c("0.3", "3", "30"), expand=expansion(0.15)) +
  scale_y_continuous(labels=scales::label_percent(), expand=expansion(mult=c(0, 0.1)),
                     breaks=c(0, 0.5, 1), limits = c(0, 1)) +
  scale_shape_manual(breaks=c("Glutamic acid", "Arginine", "Guanosine"), values=c(22, 25, 21)) +
  labs(x="Hours after <sup>15</sup>N spike", color=NULL, fill=NULL, shape=NULL,
       y="Percent metabolite\nlabeled") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    legend.position = "none"
  )

surfdeep_all_cmpd_gp <- labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(!compound_name%in%maa_names) %>%
  mutate(compound_name=str_remove_all(compound_name, "^L-")) %>%
  mutate(compound_name=str_remove(compound_name, "/beta-Glutamic acid")) %>%
  mutate(compound_class=case_when(
    compound_name%in%c("Glutamine", "Aspartic acid", "Asparagine", "Glutamic acid", "Ectoine", "Theanine?", "Citrulline", "Guanine", "Guanosine", "Arginine") ~ compound_name,
    compound_name%in%aa_names~"Other amino acids",
    compound_name%in%nucleo_names~"Other nucleic acids",
    compound_name%in%betaine_names~"Betaines",
    TRUE~"All others"
  )) %>%
  mutate(compound_class=factor(compound_class, levels=class_levels)) %>%
  summarise(avg_frac=mean(labeled_frac), .by = c(compound_class, amendment, timepoint_num, depth, tripl)) %>%
  filter(amendment!="Nit") %>%
  summarise(avg_label=mean(avg_frac), .by=c(compound_class, amendment, depth, timepoint_num)) %>%
  pivot_wider(names_from = depth, values_from=avg_label) %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Arg", "GMP"),
                          labels=c("+Ammonia", "+Arginine", "+GMP"))) %>%
  ggplot(aes(x=timepoint_num, y=Surf-Deep, color=compound_class, 
             fill=compound_class, shape=compound_class)) +
  geom_line(lwd=0.8) +
  geom_point(color="black", size=3) +
  geom_hline(yintercept = 0) +
  facet_wrap(~amendment, nrow=1) +
  scale_x_log10(breaks=c(0.3, 3, 30), labels=c("0.3", "3", "30"), expand=expansion(0.15)) +
  scale_y_continuous(labels=scales::label_percent()) +
  scale_shape_manual(breaks=setdiff(class_levels, "MAAs"), values = rep(21:25, length.out=14)) +
  labs(x="Hours after <sup>15</sup>N spike", color=NULL, fill=NULL, shape=NULL,
       y="Percent metabolite labeled at 25m\nminus percent labeled at 175m") +
  theme_bw() +
  theme(
    # axis.title.x = element_markdown(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    strip.text = element_blank(),
    strip.background = element_blank()
  )

surfdeep_all_cmpd_boxplot_gp <- labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(!compound_name%in%maa_names) %>%
  filter(amendment!="Nit") %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  summarise(labeled_frac=mean(labeled_frac), .by=c(timepoint_num, compound_name, depth, amendment)) %>%
  ggplot(aes(x=timepoint_num, y=labeled_frac, group=interaction(timepoint_num, depth), fill=depth)) +
  geom_boxplot(color="black", outlier.shape=21) +
  facet_wrap(~amendment, nrow=1) +
  scale_x_log10(breaks=c(0.3, 3, 30), labels=c("0.3", "3", "30"), expand=expansion(0.15/2)) +
  scale_y_continuous(labels = scales::label_percent(), breaks=c(0, 0.5, 1)) +
  scale_fill_manual(breaks=c("Surf", "Deep"), values=c("grey90", "grey30"), labels=c("25 meters", "175 meters")) +
  labs(x="Hours after <sup>15</sup>N spike", color=NULL, fill=NULL, shape=NULL,
       y="Percent metabolite\nlabeled") +
  theme_bw() +
  theme(
    axis.title.x = element_markdown(),
    strip.text = element_blank(),
    strip.background = element_blank()
  )

>>>>

```{r load data, eval=FALSE}
parametadata <- read_csv("metadata/parametadata.csv") %>%
  # filter(IS_type=="noIS") %>%
  filter(polarity=="pos") %>%
  filter(samp_type=="Smp" | samp_type=="Bad") %>%
  mutate(startime=factor(startime, levels=c("Morn", "Eve"))) %>%
  mutate(depth=factor(depth, levels=c("Surf", "Deep"))) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Nit", "Arg", "GMP"))) %>%
  arrange(amendment, depth, timepoint_num, startime) %>%
  mutate(timepoint=factor(timepoint, levels=c("T0", "T1", "T3", "T10", "T18", "T26", "T73")))

@@

```{r}
xamend_metab_gp <- labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, samp_type, timepoint, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(samp_type!="Bad") %>%
  filter(depth=="Surf") %>%
  mutate(compound_name=str_remove_all(compound_name, "^L-")) %>%
  mutate(compound_name=str_remove(compound_name, "/beta-Glutamic acid")) %>%
  mutate(compound_class=case_when(
    compound_name%in%c("Glutamine", "Aspartic acid", "Asparagine", "Glutamic acid", "Ectoine", "Theanine?", "Citrulline", "Guanine", "Guanosine", "Arginine") ~ compound_name,
    compound_name%in%maa_names~"MAAs",
    compound_name%in%aa_names~"Other amino acids",
    compound_name%in%nucleo_names~"Other nucleic acids",
    compound_name%in%betaine_names~"Betaines",
    TRUE~"All others"
  )) %>%
  mutate(compound_class=factor(compound_class, levels=class_levels)) %>%
  summarise(avg_frac=mean(labeled_frac), .by = c(compound_class, amendment, timepoint, tripl)) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Nit", "Arg", "GMP"),
                          labels=c("+Ammonia", "+Nitrate", "+Arginine", "+GMP"))) %>%
  ggplot() +
  geom_raster(aes(x=timepoint, y=tripl, fill=avg_frac)) +
  geom_vline(xintercept = c(1.5:4.5), color="grey80") +
  facet_nested(compound_class~amendment, scales = "free_x", switch="y") +
  scale_fill_gradientn(colours = viridis::viridis(10, alpha = seq(0, 1, length.out=10), direction = -1),
                       limits=c(0, 1), breaks=c(0, 0.25, 0.5, 0.75, 1),
                       labels=c("Fully unlabeled", "", "50% labeled", "", "Fully labeled")) +
  scale_x_discrete(expand = expansion()) +
  scale_y_discrete(expand = expansion()) +
  labs(x=NULL, y=NULL, fill=NULL) +
  theme_bw() +
  theme(panel.spacing.x=unit(0.5, "lines"), 
        panel.spacing.y = unit(0, "lines"),
        axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle=0, hjust = 1),
        strip.background.y = element_rect(fill=NA, color=NA),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "top", 
        legend.key.width = unit(1, "null")) +
  guides(
    fill=guide_colorbar(theme=theme(
      legend.frame = element_rect(color = "black"),
      legend.ticks = element_line(color="black"),
      legend.text = ggtext::element_markdown(hjust=c(0, 0, 0.5, 1, 1)))
    )
  )
ggsave("xamend_metab_gp.png",plot = xamend_metab_gp, device = "png", path = "manuscript/figures",
       width = 6.5, height = 5.5, units = "in", dpi = 300)
```

@@

```{r fig surface vs deep labeling rates}
surfdeep_demo_cmpd_gp <- labeled_fracs %>%
  mutate(compound_name=str_remove_all(compound_name, "^L-")) %>%
  mutate(compound_name=str_remove(compound_name, "/beta-Glutamic acid")) %>%
  mutate(compound_class=case_when(
    compound_name%in%c("Glutamine", "Aspartic acid", "Asparagine", "Glutamic acid", "Ectoine", "Theanine?", "Citrulline", "Guanine", "Guanosine", "Arginine") ~ compound_name,
    compound_name%in%aa_names~"Other amino acids",
    compound_name%in%nucleo_names~"Other nucleic acids",
    compound_name%in%betaine_names~"Betaines",
    TRUE~"All others"
  )) %>%
  mutate(compound_class=factor(compound_class, levels=setdiff(class_levels, "MAAs"))) %>%
  left_join(parametadata %>% distinct(shortname, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(timepoint_num!=18) %>%
  filter(compound_name=="Glutamic acid" & amendment=="Amm" |
           compound_name=="Arginine" & amendment=="Arg" |
           compound_name=="Guanosine" & amendment=="GMP") %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Arg", "GMP"),
                          labels=c("+Ammonia", "+Arginine", "+GMP"))) %>%
  select(-shortname) %>%
  summarise(mean_label=mean(labeled_frac), .by=c(compound_class, amendment, timepoint_num, depth)) %>%
  pivot_wider(names_from=depth, values_from=mean_label) %>%
  ggplot(aes(x=timepoint_num, color=compound_class, shape=compound_class)) +
  geom_segment(aes(xend=timepoint_num, y=Surf, yend=Deep), lwd=1) +
  geom_point(aes(y=Surf), fill="grey80", size=3, stroke=1) +
  geom_point(aes(y=Deep), fill="grey20", size=3, stroke=1) +
  stat_summary(aes(y=Surf, label=paste0(round(Surf-Deep, 2)*100, "%"), vjust=-sign(Surf-Deep)*1.5+0.75), 
               geom = "text", fun = mean) +
  facet_wrap(~amendment, nrow=1) +
  scale_color_discrete(drop=FALSE) +
  scale_x_log10(breaks=c(0.3, 3, 30), labels=c("0.3", "3", "30"), expand=expansion(0.15)) +
  scale_y_continuous(labels=scales::label_percent(), expand=expansion(mult=c(0, 0.1)),
                     breaks=c(0, 0.5, 1), limits = c(0, 1)) +
  scale_shape_manual(breaks=c("Glutamic acid", "Arginine", "Guanosine"), values=c(22, 25, 21)) +
  labs(x="Hours after <sup>15</sup>N spike", color=NULL, fill=NULL, shape=NULL,
       y="Percent metabolite\nlabeled") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    legend.position = "none"
  )

surfdeep_all_cmpd_gp <- labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, timepoint, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(timepoint_num!=18) %>%
  filter(!compound_name%in%maa_names) %>%
  mutate(compound_name=str_remove_all(compound_name, "^L-")) %>%
  mutate(compound_name=str_remove(compound_name, "/beta-Glutamic acid")) %>%
  mutate(compound_class=case_when(
    compound_name%in%c("Glutamine", "Aspartic acid", "Asparagine", "Glutamic acid", "Ectoine", "Theanine?", "Citrulline", "Guanine", "Guanosine", "Arginine") ~ compound_name,
    compound_name%in%aa_names~"Other amino acids",
    compound_name%in%nucleo_names~"Other nucleic acids",
    compound_name%in%betaine_names~"Betaines",
    TRUE~"All others"
  )) %>%
  mutate(compound_class=factor(compound_class, levels=class_levels)) %>%
  summarise(avg_frac=mean(labeled_frac), .by = c(compound_class, amendment, timepoint_num, depth, tripl)) %>%
  filter(amendment!="Nit") %>%
  summarise(avg_label=mean(avg_frac), .by=c(compound_class, amendment, depth, timepoint_num)) %>%
  pivot_wider(names_from = depth, values_from=avg_label) %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  mutate(amendment=factor(amendment, levels=c("Amm", "Arg", "GMP"),
                          labels=c("+Ammonia", "+Arginine", "+GMP"))) %>%
  ggplot(aes(x=timepoint_num, y=Surf-Deep, color=compound_class, 
             fill=compound_class, shape=compound_class)) +
  geom_line(lwd=0.8) +
  geom_point(color="black", size=3) +
  geom_hline(yintercept = 0) +
  facet_wrap(~amendment, nrow=1) +
  scale_x_log10(breaks=c(0.3, 3, 30), labels=c("0.3", "3", "30"), expand=expansion(0.15)) +
  scale_y_continuous(labels=scales::label_percent()) +
  scale_shape_manual(breaks=setdiff(class_levels, "MAAs"), values = rep(21:25, length.out=14)) +
  labs(x="Hours after <sup>15</sup>N spike", color=NULL, fill=NULL, shape=NULL,
       y="Percent metabolite labeled at 25m\nminus percent labeled at 175m") +
  theme_bw() +
  theme(
    # axis.title.x = element_markdown(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    strip.text = element_blank(),
    strip.background = element_blank()
  )

surfdeep_all_cmpd_boxplot_gp <- labeled_fracs %>%
  left_join(parametadata %>% distinct(shortname, timepoint_num, depth, startime, amendment, tripl)) %>%
  filter(timepoint_num!=18) %>%
  filter(!compound_name%in%maa_names) %>%
  filter(amendment!="Nit") %>%
  mutate(timepoint_num=ifelse(timepoint_num==0, 0.3, timepoint_num)) %>%
  summarise(labeled_frac=mean(labeled_frac), .by=c(timepoint_num, compound_name, depth, amendment)) %>%
  ggplot(aes(x=timepoint_num, y=labeled_frac, group=interaction(timepoint_num, depth), fill=depth)) +
  geom_boxplot(color="black", outlier.shape=21) +
  facet_wrap(~amendment, nrow=1) +
  scale_x_log10(breaks=c(0.3, 3, 30), labels=c("0.3", "3", "30"), expand=expansion(0.15/2)) +
  scale_y_continuous(labels = scales::label_percent(), breaks=c(0, 0.5, 1)) +
  scale_fill_manual(breaks=c("Surf", "Deep"), values=c("grey90", "grey30"), labels=c("25 meters", "175 meters")) +
  labs(x="Hours after <sup>15</sup>N spike", color=NULL, fill=NULL, shape=NULL,
       y="Percent metabolite\nlabeled") +
  theme_bw() +
  theme(
    axis.title.x = element_markdown(),
    strip.text = element_blank(),
    strip.background = element_blank()
  )
